\renewcommand{\abstractname}{Acknowledgements}
\begin{abstract}
Throughout the research and writing of this thesis I have received a great deal of support and assistance.
\\\\
First and foremost, I would like to express my tremendous gratitude towards my promoters, prof. dr. Frank Neven and prof. dr. Stijn Vansummeren, as well as my supervisor, dr. Lander Willem. Your guidance and expertise during and after those intensive meetings was invaluable and pushed me to get the most out of this thesis. I particularly single out Stijn, who always went the extra mile to provide the best and most detailed feedback and support, especially in the second semester.
\\\\
My fellow students, who have become lifelong friends over the years, deserve most of the credit for my academic career. I could definitely not have succeeded without your support and friendship. You were a major part in these wonderful years, not only regarding school, but also regarding all of those extracurricular evenings that I will never forget.
\\\\
Furthermore, I would like to thank my friends with whom I have spent a lot of evenings during every lockdown. Your friendship has been crucial so that I could relax and wind down during these stressy times.
\\\\
I also want to thank the person who is dearest to me, Amber, who provided unconditional support and encouragement. You have always kept me motivated and prevented me from going insane during this everlasting pandemic.
\\\\
I want to thank my parents, who gave me the possibility to study what I am passionate about and to live my own life in a student dorm.
\\\\
At last, a special thanks to my friend and physical therapist, Yenthe, whom without I physically could not have survived spending all of my time in front of my computer.
\end{abstract}